---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?
**A1.**

```{r}
umd_courses |>
  filter(str_detect(title, " Climate ")) |> filter(seats >=1) |> group_by(department) |>
  summarize(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))

```

**Q2.** Of the departments listed in A2, which one is the most interesting to you, and why?
**A2.** Art history & archaeology - climate isn't something I would associate with art. Might have more to do with archaeology?


**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?
**A3.** 
Most courses: Business + Management; School of Music; English; Teaching and Learning, Policy and Leadership; Information Studies. 
At least one syllabus posted: Business + Management; Economics; School of Music; Biological Science; Information Studies
the English department and Teaching and Learning, Policy and Leadership dropped from the top 5 (meaning they have less syllabi uploaded to their courses for the Fall 2023 term)

```{r}
umd_courses |>
  filter(term == 202308) |> 
  filter(syllabus_count >=1) |>
  group_by(department) |>
  summarise(count_class = n ()) |>
  arrange(desc(count_class))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** Wes Moore's committee spending increased over time.

```{r}

maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

head(maryland_expenses)
```
```{r}
maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```


**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5** See below. Media was the largest expense of the Moore campaign.

```{r}
maryland_expenses |>
  mutate(spending_location = case_when(
    str_detect(address, " Maryland ") ~ "In-state",
    .default = "Out of state"
  )) |>
  group_by(expense_category) |>
  summarise(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```


**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A6.**
Question: Which departments had the most courses offered with fewer than 300 seats available?
Answer: Top three are: Business and Management at 1779 classes, School of Music at 1583 classes and English at 1557 classes.

```{r}
umd_courses |>
  filter(seats <=300) |> group_by(department) |>
  summarize(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```
