---
title: "lab_06"
author: "derek willis"
date: "2023-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries
library(tidyverse)
library(janitor)
library(lubridate)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use some more 911 overdose call data to do this. The first thing we want to do is to combine multiple counties' data into a single dataframe so that we can ask some questions. First, let's combine data from Cecil, Carroll and Allegany counties into a new dataframe.

*Before* you combine them you'll need to clean up some of the column names to make the data easier to work with - make sure you have loaded the library to do that. You also need to make sure that each column has the same name and datatype (you can check the datatypes using `glimpse`). If any of them does not, you need to fix that.

```{r}
allegany_911 <- read_csv("data/allegany_911.csv") |> clean_names()
carroll_911 <- read_csv("data/carroll_911.csv") |> clean_names()
cecil_911 <- read_csv("data/cecil_911.csv") |> clean_names()

# load and combine the call data from those three counties
ACC_911_combined <- bind_rows(list(allegany_911, carroll_911, cecil_911))
View(ACC_911_combined)
```

Then we'll join some demographic data with Baltimore City 911 calls and ask some questions of that. For that, you'll need to load the Baltimore City 911 data and the ZCTA demographic data from the data folder, and join them based on the zip code. Remember, the zip codes in both dataframes must be the same datatype (and should be characters). You should start with your Baltimore City 911 dataframe in creating your join.

```{r}
# load and join the Baltimore-related data
baltimore_911 <- read_csv("data/baltimore_911.csv") |> clean_names()|> rename(zcta5n = zip_code)
maryland_zcta <- read_csv("data/maryland_zcta.csv") |> clean_names() 

baltimore_911 |> left_join(maryland_zcta, join_by("zcta5n")) |> nrow()
```

## Answer questions

Q1. Let's start with our data from Allegany, Cecil and Carroll counties. Write code to generate a dataframe that shows the total number of calls for each county. What's the order of counties from most calls to least?

A1. In order from most to least 911 calls: Carroll County at 448 calls, Cecil County at 438 calls, Allegany County at 418 calls.

```{r}
ACC_911_combined |>
  group_by(county) |>
  summarise(count_county = n()) |>
  arrange(desc(count_county))
```

Q2. In your combined dataframe that you used in A1, add a column for the month of each call, then write code to show the total number of calls per county and month. Are there any outliers or noteworthy results? Describe the general pattern of the results.

Then calculate the most calls per county and location. Which location had the most calls - where and what is that location?

A2.The summer months tend to see an increase in number of OD-related 911 calls with each county reaching peak call volume in August. Each county seemed to have different times when call volumes were lowest - generally, Cecil and Allegany counties were lowest in December/January while Carroll was lowest in April.
314 GROVE NECK RD EARLEVILLE, MD in Cecil County had the most calls (15 total). This is one of two Recovery Centers of America locations in Maryland...ironic.

```{r}
ACC_911_combined_months <- ACC_911_combined |>
  mutate(month = month(date, label = TRUE, abbr = FALSE)) |>
  group_by(county, month) |>
  summarise(total = n())

ACC_911_combined_location <- ACC_911_combined |>
  group_by(location, county) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

Q3. Let's turn to the Baltimore City data. You're interested in exploring calls in zip codes with a high percentage of children under 18. Write code to isolate calls in zip codes where the percentage of under 18 population is at least 75% and then show the zip code and population and how many calls occurred in each of those zip codes.

Where are those zip codes in the city?

Add a column to your code that calculates the number of calls per 1,000 people for those zip codes. To calculate a per capita rate, you might find this short guide handy: [https://observablehq.com/\@palewire/per-capita-calculator](https://observablehq.com/@palewire/per-capita-calculator){.uri}.

Which zip code has the highest rate of calls per 1,000 people? Find the neighborhoods in that zip code that are listed in the data - you can use group_by or distinct to do this. What are some of those neighborhoods, and what else can you tell me about the population there?

A3. The majority of the zip codes (21217, 21223, 21202, 21213, 21201, 21216) are located north and west of the downtown area. The outlier is 21205, which is located north of 1-40.
The zip code 21217 had the highest rate of calls per 1k people (20.9 calls). The zip is associated with the Upton, Madison Park, Druid Heights, and Penn North neighborhoods. Based on the Maryland_zcta file, the area is predominately black (86.4%) with over 80% of residents under the age of 18 and a median age of 34. Over two-thirds of the population live in rented housing (70%).


```{r}
baltimore_911 |> left_join(maryland_zcta, join_by("zcta5n"))
baltimore_911_demos <- baltimore_911 |> left_join(maryland_zcta) |>
  filter(pct_under_18 >= 75.0) |>
  group_by(zcta5n) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

```{r}
baltimore_911_demos <- baltimore_911 |> left_join(maryland_zcta) |>
  group_by(zcta5n, population) |>
  summarise(total = n()) |>
  mutate(calls_per_capita = (total / population) * 1000) |>
  arrange(desc(calls_per_capita))
  
```

Q4. Choose your own adventure: I want you to come up with a question you could ask of the Baltimore City data that combines both the calls and the demographic information. You must use both demographics and the overdose call data in your work.

A4:
Q. What is the rate of calls per 1,000 people in zip codes where non-Hispanic whites account for over 75% of the population? Is there anything interesting in terms of where  those zips are located in comparison to the zip codes with the highest calls per 1,000 people?
A. Non-Hispanic Whites account for at least 75% of the population in four zip codes. The highest call rate per 1,000 people in those zips is 6.5% and the lowest is 0.7%. 21211 (3.9% call rate) is on the opposite side of Druid Hill Park, where 21217 (Q3) is located.

```{r}
baltimore_911 |> left_join(maryland_zcta, join_by("zcta5n"))
baltimore_911_demos <- baltimore_911 |> left_join(maryland_zcta) |>
  filter(pct_white_non_hispanic >= 75.0) |>
  group_by(zcta5n, population) |>
  summarise(total = n()) |>
  mutate(calls_per_capita = (total / population) * 1000) |>
  arrange(desc(calls_per_capita))
```
