---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564) 

# Load in the Maryland grants and loan data

md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")
```

```{r}
library(janitor)
```

```{r}
# cleaning function
cleaned_payments <- payments |>
  clean_names()

# display the cleaned dataset
cleaned_payments
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

## Cleaned payment file below
## Answer here 
A1: 141 records found.

```{r}
pg_schools <- cleaned_payments |>
  filter(
    str_detect(vendor_name, "PRINCE GEORGES COMMUNITY COLLEGE") |
    str_detect(vendor_name, "PRINCE GEORGE'S COMMUNITY COLLEGE") |
    str_detect(vendor_name, "PRINCE GEORGE PUBLIC SCHOOLS") |
    str_detect(vendor_name, "PRINCE GEORGE'S PUBLIC SCHOOLS") |
    str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS") |
    str_detect(vendor_name, "PRINCE GEORGE'S COUNTY PUBLIC SCHOOLS") |
    str_detect(vendor_name, "PRINCE GEORGE'S COMM COLLG"))

```


Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

## Answer here
A2: 2008 had the largest total amount. 2008-2010 had billions of dollars sent to the PG County School system - they received nearly 10x less in the next highest year.

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))

```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

## Answer here
A3. Examples of vendors with similar names: WHITING TURNER CONTRACTING, VWR INTERNATIONAL INC, CANNON WASHINGTON/CANNON DESIGN INC.
To clean, we could mutate anything with "Whiting Turner" into one standard company name, mutate zip codes for same company names with differing zip codes to one standard zip.

```{r}
UMD_payments <- cleaned_payments |>
```

```{r}
UMD_payments |>
  filter(agency_name == "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND") |>
  group_by(vendor_name) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

##Answer here
A4: Spending on stem cell research was at its highest in the early 2010s (2010-2013) with a lull in the mid-2010s and a slow increase again in the late 2010s, entering the 2020s.

```{r}
cleaned_md_grants_loans |>
  #filter(category == "Grant") |>
  filter(description == "STEM CELL RESEARCH GRANT") |>
  group_by(fiscal_year) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

## Answer here
A5: A total of 18 organizations have received grant funds from MD - the majority are associated with the University of Maryland. Mutating the name for anything containing "uni" to "university of maryland" would help create one record per guarantee (except for that one UMD data center result...urg...mutate again?). While there are multiple UMD results, it's interesting that two grantee names are directly tied to our research center(s?).

```{r}
cleaned_md_grants_loans |>
  filter(zip5 == 20742) |>
  group_by(grantee) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))
```

