---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Look at the earthquakes dataset. Finish the sentence below as if you were explaining the data to someone who had not seen it before but needs to know about it.

**A1:** This dataset contains information on earthquakes from 2023-09-11 through 2023-10-11.

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2:** 9,774 records are included in the dataset. We can review location (lang/long and address), magnitude, type, date/time of occurrence.

------------------------------------------------------------------------

#### **Q3** How do I reorganize this data to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes |>
  arrange(desc(depth))

```

**A3:** The depth of the deepest earthquake was 669.9820 (kg?) and scored a magnitude of 4.20.

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
earthquakes |>
  filter(mag >6)

```

**A4:** 13 earthquakes had a magnitude larger than 6.

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
earthquakes |> filter((mag >6)&(depth <20))

```

**A5:** 6 earthquakes had a magnitude greater than 6 and a depth smaller than 20.

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 OR a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
earthquakes |> filter((mag >6)|(depth <20))
  

```

**A6:** 7,446 earthquakes either had a magnitude greater than 6 or a depth smaller than 20.

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
earthquakes |>
filter(
  str_detect(place, "Alaska"))

```

**A7:** 3,451 earthquakes occurred in Alaska.

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
earthquakes |>
  group_by(type) |>
  summarise(
    count_type = n()) |>
  arrange(desc(count_type))
```

**A8:** The two most common types are earthquake and quarry blast.

------------------------------------------------------------------------

#### **Q9** What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
earthquakes |>
  filter(type == "earthquake") |>
  summarise(
    mean_depth = mean(depth)
    )

```

**A9:** The average depth of an earthquake is 25.23739kg. Most earthquakes are caused by shifting of tectonic plates around 40kg depth. A 25kg average makes these more shallow than the average - shallow earthquakes cause more damage than deeper earthquakes as intensity of shaking is not reduced.

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  arrange(desc(hour))

```

**A10:** Mutate the data to extract the hour listed in the time column (YMD HMS).

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column showing whether an earthquake is serious or not?

```{r}
earthquakes |>
  mutate(severity = case_when(
    mag > 3 ~ "serious",
    .default = "not serious"
  ))

```

**A11:** Mutate data to create a new column that determines severity of earthquake (serious, not serious) via case_when magnitude >3.

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
earthquakes |>
  mutate(hour = hour(time)) |>
  mutate(severity = case_when(
    mag > 3 ~ "serious",
    .default = "not serious"
  )) |>
  group_by(hour, severity) |>
  summarise(count_hour = n()) |>
  arrange(desc(count_hour))

```

**A12**: 12:00 p.m. has the fewest serious earthquakes; 2 a.m. has the most serious earthquakes.

------------------------------------------------------------------------

#### **Q13** What's another question you are interested in and how would you ask it in R?

```{r} 
earthquakes |>
  mutate(severity = case_when(
    mag > 3 ~ "serious",
    .default = "not serious"
  )) |>
  group_by(net) |>
  summarise(count_net = n()) |>
  arrange(desc(count_net))

```
### **A13** Which network regions experienced the most earthquakes within the month of available data? 